Modelling circumstellar discs with 3D radiation hydrodynamics

### 2009-10-02 SPH-Torus-paper1/sphtorus_benchmark.tex
changes from meeting - th
git-svn-id: https://repository.astro.ex.ac.uk/torus/trunk@2043 25464f21-6099-402c-b30a-640f1d40d38b

Hydrodynamics calculations are used to investigate a variety of problems associated with star formation, such as the collapse of protostellar cores (e.g. \citealt{goodwin_2004}), the fragmentation of protostellar discs (e.g. \citealt{boss_1997}), and on the largest scale the gravo-turbulent collapse of a molecular cloud to form stars (e.g. \citealt{bate_2003, bate_2005}).

Sophisticated as such calculations appear, the underlying physics regarding the thermal properties of the gas is much simplified. One of the usual approximations is the use of an equation of state (eos) in determining the pressure of the gas. An isothermal eos is often a good approximation in the most rarified regions, where the material is able to radiate, and thus cool, freely. The gas may be compressed adiabatically in the highest density regions, and a barytropic eos is sometimes adopted, with the polytropic exponent switching at a critical density in an attempt to reproduce the results of one-dimensional collapse calculations that incorporate radiative-transfer \citep{masunaga_1998}. The crucial point here is that the energy dumped into the cluster from accretion onto, and gravitation contraction of, the newly formed stars is neglected, and the radiation emitted by compressed gas (the `$P\,dV$' terms) is assumed to  escape without further interaction.

@@

Meanwhile dedicated radiation transfer (RT) codes  have progressed far beyond the diffusion approximation and now take into account a wide array of microphysical processes such as polychromatic radiation transfer incorporating polarization state, multiple scattering, Mie phase matrices etc. These sophisticated RT codes have been employed to model cores (e.g. \citealt{stamatellos_2005}), discs (reviewed by \citealt{dullemond_2007}) and even entire clusters \citep{kurosawa_2004}. The principle barrier to using these codes in RHD is that of  of computation time, but the increase in CPU power and the  adoption of parallelization mean that it is now feasible to conduct a full treatment of RT along with the hydrodynamics, thereby dropping the diffusion approximation.

@@

Accurately coupling the radiative transfer component to the SPH
component requires an effective transformation from the Lagrangian
particle description of SPH to the Eulerian grid description of an AMR
grid. The procedure used for this transformation is tested using a
circumstellar disc benchmark (see
Sections~\ref{section:description_of_benchmark}
and~\ref{section:grid_generation}). The temperature distribution and spectral energy distribution
 of the benchmark disc are examined in
Section~\ref{section:seds} in order to investigate how well the
radiative transfer calculation works with the gridded density field
and to understand the effects of grid and SPH spatial resolution. A
simulation of a circumstellar disc around a T Tauri star is presented
in Section~\ref{section:sph_disc}.

@@

The accuracy of the SPH method has previously been examined in detail
(e.g. \cite{price_2008b} and references therein) so we do not attempt
to benchmark the SPH component of the code here. Likewise the  {\sc torus}
radiative transfer code has also been benchmarked elsewhere
\citep{pinte_2009}. Consequently we can be confident
in the accuracy of the SPH and radiative transfer codes individually so
we focus here on the accuracy of the mechanism by which they are
coupled. Hence the benchmark case presented here is a `static'
benchmark in which we represent an analytical density distribution
with SPH particles and test whether the radiative transfer code can
calculate accurate results using this SPH density representation.

We base our SPH disc benchmark on the axisymmetric circumstellar disc of
\cite{pascucci_2004}.  Representing a structure with rotational
symmetry using a 3D geometry allows us to validate the operation of
our method against well tested results from other codes while fully
exercising the 3D capabilities of our own system.

@@

Having investigated the procedure for setting up the radiative transfer
grid we now proceed to studying the accuracy of the results from the
radiative transfer calculation. This was carried out by comparing
temperature distributions and spectral energy distributions (SEDs)
from the benchmark disc. Temperature distribution errors provide a
clear picture of where errors in the radiative transfer calculation
are located whereas SEDs provide a direct link to an observable
quantity and are a stringent test of the accuracy of the radiative
transfer calculation.

The disc was initially modelled using $10^5$ SPH particles with an AMR
grid constructed from a mass per cell limit of $5\times10^{26}$g.  The
mass per cell limit was chosen in preference to the density contrast
limit to give the highest resolution near the central region of the
disc, which we expect to be most important for determining the
SED. SEDs were calculated using the {\sc torus} code for viewing
angles of 12.5 and 77.5 degrees (where an inclination angle of zero
degrees indicates that the disc is viewed face-on) and the results are
plotted in Fig.~\ref{fig:seds} (dashed line). Also plotted in
Fig.~\ref{fig:seds} are the benchmark results of \cite{pascucci_2004}
(solid line).
\begin{figure*}
  \includegraphics[scale=0.3]{figures/sed_013}
  \includegraphics[scale=0.3]{figures/sed_077}
  \caption{SEDs from the benchmark disc with $10^5 $ particles. SEDs
    are shown for inclination angles of 12.5 degrees (left) and 77.5
    degrees (right). The benchmark SED is plotted as a solid line and
    the {\sc torus} SED is plotted as a dashed line.}
  \label{fig:seds}
\end{figure*}
Both SEDs show significant departures from the benchmark results;
there is excess emission between 10-100 $\mu\rm{m}$ and at a viewing
angle of 77.5 degrees there is also a significant reduction in flux
below 1~$\mu\rm{m}$. The density distribution in the central region of
this disc is plotted in Fig.~\ref{fig:basic_dens} which shows that the
central 1AU gap is not represented. There are two effects in
operation; firstly the grid resolution in this region is larger than
1AU and secondly the smoothing lengths of the particles are larger
than 1AU. In order to correctly represent the central gap there needs
to be higher grid resolution around the central source and more SPH
particles are required so that the smoothing lengths are smaller.
\begin{figure}
\includegraphics[scale=0.2]{figures/basic_dens}
\caption{Density in central region of disc with $10^5$ particles. Axis
units are $10^{13}$cm.}
\label{fig:basic_dens}
\end{figure}

\subsection{Grid resolution}
Decreasing the mass per cell limit to achieve the required resolution
around the central source would result in too many cells being
generated in other regions of the disc. In order to achieve a grid
resolution of less than 1AU around the source it was necessary to
specify additional criteria for refining the grid. Three cubes of
increased resolution were added to the AMR grid with sizes
$1.5\times10^{15}$~cm, $3\times10^{14}$~cm and
$6\times10^{13}$~cm. Within each cube it was required that the cell
size be no larger than 0.1 of the cube dimension. In order to confirm
that the grid modifications provide sufficient resolution to generate
an accurate SED three calculations were performed with the central
part of the disc forced to the correct density values from the
analytical density description.  A spherical region with radius
$1.0\times10^{15}$~cm, $5.0\times10^{14}$~cm or $2.5\times10^{14}$~cm
was over written with the benchmark disc density after the initial
grid generation was performed. The resulting SEDs are plotted in
Fig.~\ref{fig:sed_forced}. The solid line shows the benchmark result
and the other lines show SEDs for forcing regions of radius
$1.0\times10^{15}$~cm (long dashed line), $5.0\times10^{14}$~cm (short
dashed line) and $2.5\times10^{14}$~cm (dotted line).
\begin{figure*}
  \includegraphics[scale=0.3]{figures/seds_forced_13.pdf}
  \includegraphics[scale=0.3]{figures/seds_forced_77.pdf}
  \caption{SEDs from SPH benchmark disc for inclination angles of 12.5
    degrees and 77.5 degrees. The benchmark SED is plotted as a solid
    line. The {\sc torus} density distribution is forced to the correct
    value with a radius of $1.0\times10^{15}$~cm (long dashed line), $5.0\times10^{14}$~cm
(short dashed line) and $2.5\times10^{14}$~cm (dotted line).} 
  \label{fig:sed_forced}
\end{figure*}
With the correct density forced out to a radius of
$1.0\times10^{15}$~cm the SED is close to the benchmark result for
both viewing angles, however if the forced region is reduced in size
there is a significant discrepancy relative to the benchmark SED. This
confirms that the grid now has enough resolution to generate an
accurate SED but also indicates that using $10^5$ particles does not
give a sufficiently accurate density distribution within the central
$1.0\times10^{15}$~cm of the disc.

\subsection{Number of particles}
Further calculations were performed using different numbers of SPH
particles and the enhanced grid resolution described above in order to
determine the impact of the number of SPH particles on the accuracy of
the radiative transfer calculation. The density distribution in the
central region of the disc is plotted in
Fig.~\ref{fig:n_part_central_dens} in order to show how the
representation improves as the number of particles is increased.
\begin{figure*}
  \subfigure[$10^5$ particles]{\includegraphics[scale=0.22]{figures/dens_1e5}}
  \subfigure[$10^6$ particles]{\includegraphics[scale=0.22]{figures/dens_1e6}}
  \subfigure[$10^7$ particles]{\includegraphics[scale=0.22]{figures/dens_1e7}}
  \subfigure[$10^8$ particles]{\includegraphics[scale=0.22]{figures/dens_1e8}}
  \caption{Central density on AMR grid derived from different numbers
    of SPH particles. The axis units are $10^{13}$~cm.} 
  \label{fig:n_part_central_dens}
\end{figure*}
With $10^5$ and $10^6$ particles a central gap is present but is too
large. With $10^7$ particles there is more material close in to the
source but the gap itself is not represented and the source is
obscured. Only with $10^8$ particles is a gap of approximately the
correct size present.  The representation of the central 1~AU gap is
significantly affected by the number of particles used to represent
the disc, with a large number of particles required to achieve an
accurate density distribution in this spatially small region.

Figure~\ref{fig:n_part_tem_diff} shows fractional errors in the
temperature distribution for different numbers of
particles. {\sc torus} calculates the temperature distribution
using a high resolution 2D geometry, in order to determine the correct
temperature as a function of cylindrical polar $r$ and $z$
co-ordinates. This is then compared to the temperature from the SPH
disc for each SPH particle. Positive values of the fractional error
indicate that the SPH disc is hotter than the expected value.
\begin{figure*}
  \subfigure[$10^5$ particles]{\includegraphics[scale=0.3]{figures/tdiff_1e5}}
  \subfigure[$10^6$ particles]{\includegraphics[scale=0.3]{figures/tdiff_1e6}}
  \subfigure[$10^7$ particles]{\includegraphics[scale=0.3]{figures/tdiff_1e7}}
  \subfigure[$10^8$ particles]{\includegraphics[scale=0.3]{figures/tdiff_1e8}}
  \caption{Fractional temperature error for discs represented by
    different numbers of SPH particles. Axis units are AU.} 
  \label{fig:n_part_tem_diff}
\end{figure*}
When using $10^5$ particles to represent the disc there is a too much
heating in the disc mid plane and insufficient heating further out of
the mid plane. The effect is reduced as the number of particles is
increased and reaches a very low level with $10^8$ particles.

@@

SEDs are plotted in Fig.~\ref{fig:n_part_sed_013} for a viewing angle
of 12.5 degrees and Fig.~\ref{fig:n_part_sed_077} for a viewing angle
of 77.5 degrees. At the shortest wavelengths the 12.5 degrees
inclination angle SED is dominated by emission from the central
star. This region of the SED is well represented with $10^5$ and
$10^8$ particles but is less well represented with $10^6$ or $10^7$
particles where the central star is obscured. At the longest
wavelengths the disc is optically thin and the SEDs will be well
modelled provided the total disc mass is accurately represented. In
all four cases the total disc mass is well represented and the long
wavelength SED matches the benchmark results. At intermediate
wavelengths the SED is sensitive to the details of the disc
representation and it is clear that accurately modelling the SED from
this disc is challenging, even with $10^8$ particles. 
\begin{figure*}
  \subfigure[$10^5$ particles]{\includegraphics[scale=0.3]{figures/sed_013_1e5}}
  \subfigure[$10^6$ particles]{\includegraphics[scale=0.3]{figures/sed_013_1e6}}
  \subfigure[$10^7$ particles]{\includegraphics[scale=0.3]{figures/sed_013_1e7}}
  \subfigure[$10^8$ particles]{\includegraphics[scale=0.3]{figures/sed_013_1e8}}
  \caption{SEDs at a 12.5 degree inclination angle for discs represented by different numbers of SPH
    particles. The benchmark result is plotted as a solid line and the
  modelled SED is plotted as a dashed line. } 
  \label{fig:n_part_sed_013}
\end{figure*}
\begin{figure*}
  \subfigure[$10^5$ particles]{\includegraphics[scale=0.3]{figures/sed_077_1e5}}
  \subfigure[$10^6$ particles]{\includegraphics[scale=0.3]{figures/sed_077_1e6}}
  \subfigure[$10^7$ particles]{\includegraphics[scale=0.3]{figures/sed_077_1e7}}
  \subfigure[$10^8$ particles]{\includegraphics[scale=0.3]{figures/sed_077_1e8}}
  \caption{SEDs at a 77.5 degree inclination angle for discs represented by different numbers of SPH
    particles. The benchmark result is plotted as a solid line and the
  modelled SED is plotted as a dashed line. } 
  \label{fig:n_part_sed_077}
\end{figure*}

@@

@@

The disc is then evolved using the combined SPH and radiative transfer
code. In this configuration the internal energies of the SPH particles
are set using the temperatures calculated by the radiative transfer
code. The SPH part of the code does not change the particle
temperatures and they retain the temperature assigned by the radiative
transfer code until the next radiation time step. The temperatures
calculated by the radiative transfer code are equilibrium
temperatures, as it is assumed that the time scale for radiative
equilibrium is short compared to the dynamical time scale of the disc
(i.e. the ``prompt escape'' regime described by \cite{nayakshin_2009}).
A radiative transfer calculation is performed after every four
hydrodynamics time steps (equivalent to every 2.4 years). The frequency
with which the radiative transfer is calculated is determined by the
sound crossing time in the inner regions of the disc as this is the
time scale over which the scale height in this region will vary and
effect the transfer of radiation into the outer parts of the disc.

A cylindrical polar grid, with adaptive $r$ and $z$ cells and 16
azimuthal cells, was used for the radiative transfer step. The AMR grid
was constructed using a maximum mass per cell of
$3.0\times10^{25}~\rm{g}$ and a density splitting condition of
$f_{\rm{split}}=0.1$. The density in a cylindrical region around the
source was reduced to a very low value to ensure that the disc did not
completely obscure the source. A real circumstellar disc will be
truncated by the stellar magnetosphere so there is a physical
justification for enforcing a low density region around the star. The
radius of this region is $R_{\rm{gap}}=1.0~\rm{AU}$ which corresponds
to the accretion radius of the SPH sink particle. Two extra levels of
grid refinement were added in a box around the central source to
ensure that the grid had enough resolution to allow the gap around the
source to be resolved. The first box of extra refinement has size
$10^{14}~\rm{cm}$ with cells no larger than $10^{13}~\rm{cm}$ and the
second box has size $4\times10^{13}~\rm{cm}$ with cells no larger than
$4\times10^{12}~\rm{cm}$.

Monte-Carlo radiative transfer calculations can be computationally
demanding so the calculation was parallelised using MPI (Message
Passing Interface) libraries. Parallelising code in this way allows
for parallel computation in both the radiation and hydrodynamics steps
and can be applied to a shared memory or a distributed memory computer
architecture. Evolving the disc using both the radiation and
hydrodynamics codes took a total wall time of 560 hours using 128 MPI
processes, of which 440 hours was spent in radiation calculations. The
radiation calculations take approximately 4~times as long as the
hydrodynamics calculations so the combined code is significantly
slower than the SPH code alone. However the calculation is still
feasible and we expect the combined code to be used in applications
where the increased accuracy of the Monte-Carlo radiation calculation,
over faster alternatives, is of importance.

@@

Transient oscillations, as the disc adjusts towards equilibrium, are
observed whether radiative transfer is included or not. They are
vertical motions which decay more rapidly at smaller radii and hence
give the appearance of propagating outwards, even though the motion is
vertical. Other disc models using radiation hydrodynamics have found
oscillatory solutions with inwards propagating waves
(e.g. \cite{min_2009}) which are fundamentally different to the
transient effects seen in this case. However it should be noted that
\cite{min_2009} only found an oscillatory solution in their most
optically thick disc, which is significantly more massive than our
model disc, and may represent a different regime to that considered
here.

>>>>

Hydrodynamics calculations are used in investigations of  a variety of problems associated with star formation, such as the collapse of protostellar cores (e.g. \citealt{goodwin_2004}), the fragmentation of protostellar discs (e.g. \citealt{boss_1997}), and on the largest scale the gravo-turbulent collapse of a molecular cloud to form stars (e.g. \citealt{bate_2003, bate_2005}).

Sophisticated as such calculations appear, the underlying physics regarding the thermal properties of the gas is much simplified. One of the usual approximations is the use of an equation of state (eos) in determining the pressure of the gas. An isothermal eos is often a good approximation in the most rarified regions, where the material is able to radiate, and thus cool, freely. The gas may be compressed adiabatically in the highest density regions, and a barytropic eos is sometimes adopted, with the polytropic exponent switching at a critical density in an attempt to reproduce the results of one-dimensional collapse calculations that incorporate radiative-transfer \citep{masunaga_1998}. The crucial point here is that the energy dumped into the cluster from accretion onto, and gravitation contraction of, the newly formed stars is neglected, and the radiation emitted by compressed gas (the `$P\,dV$' term) is assumed to  escape without further interaction.

@@

Meanwhile dedicated radiation transfer (RT) codes  have progressed far beyond the diffusion approximation and now take into account a wide array of processes such as polychromatic radiation transfer incorporating the radiation's polarization state, multiple scattering, Mie phase matrices etc. These sophisticated RT codes have been employed to model cores (e.g. \citealt{stamatellos_2005}), discs (reviewed by \citealt{dullemond_2007}) and even entire clusters \citep{kurosawa_2004}. The principle barrier to using these codes in RHD is that of  of computation time, but the increase in CPU power and the  adoption of parallelization mean that it is now feasible to conduct a full treatment of RT along with the hydrodynamics, thereby dropping the diffusion approximation.

@@

Accurately coupling the radiative transfer component to the SPH
component requires an effective transformation from the Lagrangian
particle description of SPH to the Eulerian grid description of an AMR
grid. The procedure used for this transformation is tested using a
circumstellar disc benchmark (see
Sections~\ref{section:description_of_benchmark}
and~\ref{section:grid_generation}). The temperature distribution and spectral energy distribution
 of the benchmark disc are examined in
Section~\ref{section:seds} in order to investigate how well the
radiative transfer calculation works with the gridded density field
and to understand the effects of grid and SPH spatial resolution. Finally the hybrid code is 
tested under realistic conditions via a simulation of a circumstellar disc around a 
T Tauri star (Section~\ref{section:sph_disc}).

@@

The accuracy of the SPH method has previously been examined in detail
(e.g. \cite{price_2008b} and references therein) so we do not attempt
to benchmark the SPH component of the code here. Likewise the  {\sc torus}
radiative transfer code has also been benchmarked elsewhere
\citep{pinte_2009}. Consequently we can be confident
in the accuracy of the SPH and radiative transfer codes individually and we instead
 focus on the accuracy of the mechanism by which they are
coupled. The benchmark case presented here is therefore a `static'
benchmark in which we represent an analytical density distribution
with SPH particles and test whether the radiative transfer code can
calculate accurate results using this SPH density representation.

We base our SPH disc benchmark on the axisymmetric circumstellar disc of
\cite{pascucci_2004}.  Representing a structure with rotational
symmetry using a 3D geometry allows us to validate the operation of
our method against well tested results from other codes while fully
exercising the 3D capabilities of our own system. The geometry is a stringent test of the gridding method, since it contains a large range in linear scales that need to be resolved, and sharp density gradients (in particular the disc inner edge).

@@

Having investigated the procedure for setting up the radiative transfer
grid we now proceed to studying the accuracy of the results from the
radiative transfer calculation. This was carried out by comparing
temperature distributions and spectral energy distributions (SEDs)
from the benchmark disc. Temperature distribution errors provide a
clear picture of where errors in the radiative transfer calculation
are located whereas SEDs provide a direct link to an observable
quantity and are a stringent test of the accuracy of the radiative
transfer calculation. The key problem here is that the resolution necessary for adequately
describing the fluid flow is not necessarily the same as required for accurately 
modelling the radiation field.

The disc was initially modelled using $10^5$ SPH particles with an AMR
grid constructed from a mass per cell limit of $5\times10^{26}$g.  
This number of SPH particles would normally be considered as sufficient to adequately represent a disc of this mass in a purely hydrodynamical calculation. The mass per cell limit was chosen in preference to the density contrast
limit to give the highest resolution near the central region of the
disc, which we expect to be most important for determining the
temperature of the disc at the inner edge and the SED.

In figure xxx we plot the temperature distribution at radiative equilibrium for the benchmark disc. 
{\sc torus} calculates the temperature distribution
using a high resolution 2D geometry, in order to determine the correct
temperature as a function of cylindrical polar $r$ and $z$
co-ordinates. This is then compared to the temperature from the SPH
disc for each SPH particle. Positive values of the fractional error
indicate that the SPH disc is hotter than the expected value. The temperatures show deviations of $\pm$ 50\% (XXXX CHECK) from the benchmark values, a deviation that is purely driven by the resolution of the density field derived from the SPH particles, since a `{\sc torus} only' calculation, where the AMR grid is refined on optical depth criteria and the cell densities are found from the analytical description of the density structure (equation xxxx), matches the benchmark temperatures to within 2\%. The systematic differences in the temperature distribution are due to an increased vertical transport of radiation from the surface layers of the disc to the mid-plane due to a lack of spatial resolution, leading to errors in the calculation of the specific intensities (particularly in the inner disc region).  These systematic differences are more readily apparent in the emergent SEDs, which are a more sensitive probe of the accuracy of the RT calculation than  the temperature distribution.

SEDs were calculated using the {\sc torus} code for viewing
angles of 12.5 and 77.5 degrees (where an inclination angle of zero
degrees indicates that the disc is viewed face-on) and the results are
plotted in Fig.~\ref{fig:seds} (dashed line). Also plotted in
Fig.~\ref{fig:seds} are the benchmark results of \cite{pascucci_2004}
(solid line).
\begin{figure*}
  \includegraphics[scale=0.3]{figures/sed_013}
  \includegraphics[scale=0.3]{figures/sed_077}
  \caption{SEDs from the benchmark disc with $10^5 $ particles. SEDs
    are shown for inclination angles of 12.5 degrees (left) and 77.5
    degrees (right). The benchmark SED is plotted as a solid line and
    the {\sc torus} SED is plotted as a dashed line.}
  \label{fig:seds}
\end{figure*}
Both SEDs show significant departures from the benchmark results;
there is excess emission between 10-100 $\mu\rm{m}$ and at a viewing
angle of 77.5 degrees there is also a significant reduction in flux
below 1~$\mu\rm{m}$. Both these discrepancies are attributable to a lack of resolution within a few au of the central object. The density distribution in the central region of
this disc is plotted in Fig.~\ref{fig:basic_dens} which shows that the
central 1\,au gap is not represented. There are two effects in
operation; firstly the grid resolution in this region is larger than
1\,au  and secondly the smoothing lengths of the particles are larger
than 1\,au. In order to correctly represent the central gap there needs
to be higher grid resolution around the central source and more SPH
particles are required so that the smoothing lengths are smaller.
\begin{figure}
\includegraphics[scale=0.2]{figures/basic_dens}
\caption{Density in central region of disc with $10^5$ particles. Axis
units are $10^{13}$cm.}
\label{fig:basic_dens}
\end{figure}

\subsection{Grid resolution}
Decreasing the mass per cell limit to achieve the required resolution
around the central source would result in too many cells being
generated in other regions of the disc. In order to achieve a grid
resolution of less than 1\,au around the source it was necessary to
specify additional criteria for refining the grid. Three cubes of
increased resolution were added to the AMR grid with sizes
$1.5\times10^{15}$~cm, $3\times10^{14}$~cm and
$6\times10^{13}$~cm. Within each cube it was required that the cell
size be no larger than 0.1 of the cube dimension. In order to confirm
that the grid modifications provide sufficient resolution to generate
an accurate SED three calculations were performed with the central
part of the disc forced to the correct density values from the
analytical density description.  A spherical region with radius
$1.0\times10^{15}$~cm, $5.0\times10^{14}$~cm or $2.5\times10^{14}$~cm
was over written with the benchmark disc density after the initial
grid generation was performed. The resulting SEDs are plotted in
Fig.~\ref{fig:sed_forced}. The solid line shows the benchmark result
and the other lines show SEDs for forcing regions of radius
$1.0\times10^{15}$~cm (long dashed line), $5.0\times10^{14}$~cm (short
dashed line) and $2.5\times10^{14}$~cm (dotted line).
\begin{figure*}
  \includegraphics[scale=0.3]{figures/seds_forced_13.pdf}
  \includegraphics[scale=0.3]{figures/seds_forced_77.pdf}
  \caption{SEDs from SPH benchmark disc for inclination angles of 12.5
    degrees and 77.5 degrees. The benchmark SED is plotted as a solid
    line. The {\sc torus} density distribution is forced to the correct
    value with a radius of $1.0\times10^{15}$~cm (long dashed line), $5.0\times10^{14}$~cm
(short dashed line) and $2.5\times10^{14}$~cm (dotted line).} 
  \label{fig:sed_forced}
\end{figure*}
With the correct density forced out to a radius of
$1.0\times10^{15}$~cm the SED is close to the benchmark result for
both viewing angles, however if the forced region is reduced in size
there is a significant discrepancy relative to the benchmark SED. This
confirms that the grid now has enough resolution to generate an
accurate SED but also indicates that using $10^5$ particles does not
give a sufficiently accurate density distribution within the central
$1.0\times10^{15}$~cm of the disc.

\subsection{Number of particles}
Further calculations were performed using different numbers of SPH
particles and the enhanced grid resolution described above in order to
determine the impact of the number of SPH particles on the accuracy of
the radiative transfer calculation. The density distribution in the
central region of the disc is plotted in
Fig.~\ref{fig:n_part_central_dens} in order to show how the
representation improves as the number of particles is increased.
\begin{figure*}
  \subfigure[$10^5$ particles]{\includegraphics[scale=0.22]{figures/dens_1e5}}
  \subfigure[$10^6$ particles]{\includegraphics[scale=0.22]{figures/dens_1e6}}
  \subfigure[$10^7$ particles]{\includegraphics[scale=0.22]{figures/dens_1e7}}
  \subfigure[$10^8$ particles]{\includegraphics[scale=0.22]{figures/dens_1e8}}
  \caption{Central density on AMR grid derived from different numbers
    of SPH particles. The axis units are $10^{13}$~cm.} 
  \label{fig:n_part_central_dens}
\end{figure*}
The effective resolution of the AMR mesh is dependent on two factors; the probability of having sufficient particles to adequately sample the small volume of the disc inner-edge, and the particle smoothing length (which determines the effective spatial resolution of the SPH density representation).
With $10^5$ and $10^6$ particles a central gap is present but is too
large. With $10^7$ particles there is more material close in to the
source but the gap itself is not represented (because the smoothing length of the particles near the disc inner-edge is too large) and the source is
obscured. Only with $10^8$ particles is a gap of approximately the
correct size present.  The representation of the central 1\,au gap is
significantly affected by the number of particles used to represent
the disc, with a large number of particles required to achieve an
accurate density distribution in this spatially small region.

Figure~\ref{fig:n_part_tem_diff} shows fractional errors in the
temperature distribution for different numbers of
particles. 
\begin{figure*}
  \subfigure[$10^5$ particles]{\includegraphics[scale=0.3]{figures/tdiff_1e5}}
  \subfigure[$10^6$ particles]{\includegraphics[scale=0.3]{figures/tdiff_1e6}}
  \subfigure[$10^7$ particles]{\includegraphics[scale=0.3]{figures/tdiff_1e7}}
  \subfigure[$10^8$ particles]{\includegraphics[scale=0.3]{figures/tdiff_1e8}}
  \caption{Fractional temperature error for discs represented by
    different numbers of SPH particles. Axis units are au.} 
  \label{fig:n_part_tem_diff}
\end{figure*}
When using $10^5$ particles to represent the disc there is a too much
heating in the disc mid plane and insufficient heating further out of
the mid plane. The effect is reduced as the number of particles is
increased and reaches a very low level with $10^8$ particles, at which stage the
temperature distribution agrees with the benchmark to within $\sim$10\%.

@@

SEDs are plotted in Fig.~\ref{fig:n_part_sed_013} for a viewing angle
of 12.5 degrees and Fig.~\ref{fig:n_part_sed_077} for a viewing angle
of 77.5 degrees. At the shortest wavelengths the 12.5 degrees
inclination angle SED is dominated by the photosphere of the central
star. This region of the SED is well represented with $10^5$ and
$10^8$ particles but is of course less well represented with $10^6$ or $10^7$
particles where the central star is obscured. At the longest
wavelengths the disc is optically thin and the SEDs will be well
modelled provided the total disc mass is accurately represented. In
all four cases the total disc mass is well represented (XXX cite section) and the long
wavelength SED matches the benchmark results. At intermediate
wavelengths the SED is sensitive to the details of the disc
representation and it is clear that accurately modelling the SED from
this disc is challenging, even with $10^8$ particles. (XXXX need to discuss this in the conclusion).
\begin{figure*}
  \subfigure[$10^5$ particles]{\includegraphics[scale=0.3]{figures/sed_013_1e5}}
  \subfigure[$10^6$ particles]{\includegraphics[scale=0.3]{figures/sed_013_1e6}}
  \subfigure[$10^7$ particles]{\includegraphics[scale=0.3]{figures/sed_013_1e7}}
  \subfigure[$10^8$ particles]{\includegraphics[scale=0.3]{figures/sed_013_1e8}}
  \caption{SEDs at a 12.5 degree inclination angle for discs represented by different numbers of SPH
    particles. The benchmark result is plotted as a solid line and the
  modelled SED is plotted as a dashed line. } 
  \label{fig:n_part_sed_013}
\end{figure*}
\begin{figure*}
  \subfigure[$10^5$ particles]{\includegraphics[scale=0.3]{figures/sed_077_1e5}}
  \subfigure[$10^6$ particles]{\includegraphics[scale=0.3]{figures/sed_077_1e6}}
  \subfigure[$10^7$ particles]{\includegraphics[scale=0.3]{figures/sed_077_1e7}}
  \subfigure[$10^8$ particles]{\includegraphics[scale=0.3]{figures/sed_077_1e8}}
  \caption{SEDs at a 77.5 degree inclination angle for discs represented by different numbers of SPH
    particles. The benchmark result is plotted as a solid line and the
  modelled SED is plotted as a dashed line. } 
  \label{fig:n_part_sed_077}
\end{figure*}

@@

The above analysis indicates that the hybrid code is capable of successfully passing the density distribution from the hydrodynamics step to the radiative-transfer code, and that accurate temperatures assuming radiative equilibrium can be passed back to the SPH code in order to update the gas pressure and perform the next hydrodynamical step. In order to test the feasibility of performing a radiation-hydrodynamics calculation using the hybrid code we adopt a `realistic' test problem, that of a circumstellar disc around a classical T Tauri star. The mid-plane optical depths here are higher than that of the Pascucci benchmark, providing a more demanding challenge both in terms of computation time and required numerical accuracy.

@@

The disc is then evolved using the combined SPH and radiative transfer
code. In this configuration the internal energies of the SPH particles
are set using the temperatures calculated by the radiative transfer
code. The SPH part of the code does not change the particle
temperatures and they retain the temperature assigned by the radiative
transfer code until the next radiation time step. The temperatures
calculated by the radiative transfer code are equilibrium
temperatures, as it is assumed that the time scale for radiative
equilibrium is short compared to the dynamical time scale of the disc
(i.e. the ``prompt escape'' regime described by \cite{nayakshin_2009}).

The frequency with which the radiative transfer is calculated is determined by the
sound crossing time in the inner regions of the disc as this is the
time scale over which the scale height in this region will vary and
effect the transfer of radiation into the outer parts of the disc. (XXX put numbers
here to justify the 1 every 4). A radiative transfer calculation is performed after every four
hydrodynamics time steps (equivalent to every 2.4 years).

A cylindrical polar grid, with adaptive $r$ and $z$ cells was used for the radiative transfer step. 
The cells were split in the azimuthal direction to give a cell spacing of 22.5 degrees. This azimuthal splitting allows for the existence of non-axisymmetric instabilities (such as radiatively driven warps) which may occur due to either physical or numerical effects, while ensuring the calculation remains tractable.  The AMR grid
was constructed using a maximum mass per cell of
$3.0\times10^{25}~\rm{g}$ and a density splitting condition of
$f_{\rm{split}}=0.1$. The density in a cylindrical region around the
source was reduced to a very low value to ensure that the disc did not
completely obscure the source. A real circumstellar disc will be
truncated by the stellar magnetosphere so there is a physical
justification for enforcing a low density region around the star. The
radius of this region is $R_{\rm{gap}}=1.0~\rm{AU}$ which corresponds
to the accretion radius of the SPH sink particle. Two extra levels of
grid refinement were added in a box around the central source to
ensure that the grid had enough resolution to allow the gap around the
source to be resolved. The first box of extra refinement has size
$10^{14}~\rm{cm}$ with cells no larger than $10^{13}~\rm{cm}$ and the
second box has size $4\times10^{13}~\rm{cm}$ with cells no larger than
$4\times10^{12}~\rm{cm}$.

Monte-Carlo radiative transfer calculations can be computationally
demanding so the calculation was parallelised using MPI (Message
Passing Interface) libraries. Parallelising code in this way allows
for parallel computation in both the radiation and hydrodynamics steps
and can be applied to a shared memory or a distributed memory computer
architecture. Evolving the disc using both the radiation and
hydrodynamics codes took a total wall time of 560 hours using 128 MPI
processes, of which 440 hours was spent in radiation calculations. The
radiation calculations take approximately 4~times as long as the
hydrodynamics calculations so the combined code is significantly
slower than the SPH code alone. However the calculation is still
feasible and we expect the combined code only to be used in applications
where the increased accuracy of the Monte-Carlo radiation calculation,
over faster alternatives such as flux-limited diffusion, is of importance.

@@

Transient oscillations, as the disc adjusts towards equilibrium, are
observed whether radiative transfer is included or not. They are
vertical motions which decay more rapidly at smaller radii and hence
give the appearance of propagating outwards, even though the motion is
vertical. Other disc models using radiation hydrodynamics have found
oscillatory solutions with inwards propagating waves
(e.g. \cite{min_2009}) which are fundamentally different to the
transient effects seen in this case. However it should be noted that
\cite{min_2009} only found an oscillatory solution in their most
optically thick disc, which is significantly more massive than our
model disc, and may represent a different regime to that considered
here. We note that the disc retains its azimuthal symmetry throughout 
(XXXX Check!) the simulation and we find no evidence for warping instabilities.
